---
title: "Community-Level Health Insurance Coverage by Race and Foreign-Born Status"
format: docx
editor: visual
bibliography: health-insurance-citations.bib
csl: apa_6th.csl
  # this csl file creating using a copy of text from this file
  # https://github.com/citation-style-language/styles/blob/master/apa-6th-edition.csl
  # under the Citation Style Language open-source repo: https://github.com/citation-style-language
number-sections: true
execute: 
  echo: false
  message: false
  error: false
---

```{r}
#| label: load necessary packages and values
packages_list <- c("tidycensus", # for pulling census data
                   "tigris",     # for pulling census shape files
                   "sf",         # for handling shape files
                   "ggplot2",    # for plotting 
                   "scales",     # for scaling values
                   "dplyr",      # for data manipulation
                   "tidyr",      # for data reshaping
                   "broom",      # for easy processing of regression output
                   "flextable"   # for display tables
                   )

# Quietly load each package, and install them if not already available
for (p in packages_list) {
  if (!p %in% installed.packages()[, "Package"]) install.packages(p)
  suppressPackageStartupMessages(library(p, 
                                         character.only = TRUE,
                                         verbose = FALSE))
}
```

```{r}
#| label: pull and instatiate census key
if (file.exists("profile.R")) {
  source("profile.R")
  census_api_key(census_key)
} else {
  message("A `profile.R` script was not found in the working directory. You need to
          create that according to the instructions of the README for this repository
          in order to run this code.\n\nSee: https://github.com/nsmader/git-quarto-demo")  
  stop()
}
```

```{r}
#| label: set options

my_state  <- "Georgia"
my_county <- "Fulton" 
  # For use in the text below, provide only the name of the county, and omit
  # "County". E.g., to specify "Cook County", just write "Cook"
acs_year <- 2021
  # This is the year of American Community Survey data to draw on
```

# Introduction

Health Insurance is good, but there are disparities. This exercise looks at data from the `r acs_year` release of the American Community Survey for `r my_county` in `r my_state` to examine spatial trends in lack of health care coverage for individuals aged 35-44, and patterns of association with Black identity and foreign born status.

Consider this paper to be a counterpart to @thamer_health_1997.

# Data

We pull data from the American Community Survey 5-Year release on:

* Health Insurance Coverage By Sex And Age ([Table B27001](https://www.socialexplorer.com/data/ACS2021_5yr/metadata/?ds=ACS21_5yr&table=B27010))
* Hispanic Or Latino Origin By Race ([Table B03002](https://www.socialexplorer.com/data/ACS2021_5yr/metadata/?ds=ACS21_5yr&table=B03002))
* Sex By Age For the Foreign-Born Population ([Table B05013](https://www.socialexplorer.com/data/ACS2021_5yr/metadata/?ds=ACS21_5yr&table=B05013))

```{r}
#| label: pull census data

# For reference, see https://walker-data.com/tidycensus/articles/basic-usage.html

acs <- 
  get_acs(state     = my_state, 
          county    = my_county,
          geography = "tract", 
          variables = c(pop_35to44_male         = "B27001_015",
                        pop_35to44_noins_male   = "B27001_017",
                        pop_35to44_female       = "B27001_043",
                        pop_35to44_noins_female = "B27001_045",
                        
                        pop_35to39_foreignborn_male   = "B05013_010",
                        pop_40to44_foreignborn_male   = "B05013_011",
                        pop_35to39_foreignborn_female = "B05013_029",
                        pop_40to44_foreignborn_female = "B05013_030",

                        pop           = "B03002_001",
                        pop_blacknonh = "B03002_004"),
          year = acs_year) |>
  select(-moe) |>
  pivot_wider(names_from = "variable",
              values_from = "estimate") |>
  mutate(pop_35to44       = pop_35to44_male       + pop_35to44_female,
         pop_35to44_noins = pop_35to44_noins_male + pop_35to44_noins_female,
         pct_35to44_noins = pop_35to44_noins / pop_35to44,
         
         pop_35to44_foreignborn = 
           pop_35to39_foreignborn_male +
           pop_35to39_foreignborn_female +
           pop_40to44_foreignborn_male +
           pop_40to44_foreignborn_female,
         pct_35to44_foreignborn = pop_35to44_foreignborn / pop_35to44,
         
         pct_blacknonh = pop_blacknonh / pop) |>
  select(GEOID, pop_35to44, pct_35to44_noins, pct_35to44_foreignborn, pct_blacknonh)
```

```{r}
#| label: pull census shape files and merge in 

my_tracts <- 
  tracts(state = my_state, 
         county = my_county,
         cb = TRUE,
         progress_bar = FALSE)

tract_data <- 
  merge(my_tracts,
        acs,
        by = "GEOID")
```


# Spatial Patterns in `r my_county` County {#sec-maps}

```{r}
#| label: develop plotting function

map_field <- function(field, vir_color = "viridis") {
  ggplot(tract_data) +
  geom_sf(aes(fill = get(field)),
          color = NA) +
  scale_fill_viridis_c(name = "",
                       labels = percent,
                       option = vir_color) +
  theme_void()
}
```

## Uninsured Status

Figure @fig-uninsured shows rates of uninsured populations age 35-44 in census tracts across `r my_county`.

```{r fig.cap = paste0("Percent Uninsured in ", my_county, " County, Age 35-44")}
#| label: fig-uninsured
#| fig-width: 4
map_field("pct_35to44_noins")
```

## Race and Foreign-Born Identities

Next we examine other community characteristics.

```{r fig.cap = paste0("Demographic Characteristics of ", my_county, " County")}
#| label: fig-demographics
#| fig-cap-location: top
#| layout-nrow: 1
#| fig-subcap: 
#|   - "Percent Black Non-Hispanic"
#|   - "Percent Foreign Born, Age 35-44"

map_field("pct_blacknonh", 
          vir_color = "magma")

map_field("pct_35to44_foreignborn",
          vir_color = "magma")
```

Maps @fig-demographics-1 and @fig-demographics-2 show distinctive patterns.

# Statistical Associations

We use statistical regression methods to examine the associations between the rate of uninsured individuals and the community characteristics displayed in section @sec-maps.

The equation for Ordinary Least Squares is:

$$\hat{\beta}=(X^{\prime}X)^{-1}X^{\prime}Y$$ {#eq-ols}

@eq-ols is a classic.

```{r}
#| label: run regression
reg <- lm(pct_35to44_noins ~ pct_35to44_foreignborn + pct_blacknonh, 
          data = acs,
          weights = pop_35to44)

coeffs <- tidy(reg)
```

Table 1 shows that community composition in both foreign-born status and black non-Hispanic identity are significantly related to community uninsured status.

```{r}
#| label: tbl-reg
#| tbl-cap: "Associations of Rate of Uninsured vs Community Characteristics"

coeffs |>
  mutate(term = recode(term, 
                       "pct_35to44_foreignborn" = "% Foreign Born, Age 35-44",
                       "pct_blacknonh"          = "% Black Non-Hispanic")) |>
  flextable() |> 
  set_header_labels(term = "",
                    estimate = "Estimate",
                    std.error = "Std. Err.",
                    statistic = "t-stat",
                    p.value    = "p-value") |>
  colformat_double(digits = 3) |>
  width(j = "term",
        width = 2)
```


# References

::: {#refs}
:::

